Canonical rings of rational divisors on projective space and Hirzebruch surfaces

### 2015-07-11 surfaces-write-up/surface-write-up.tex

Write
\begin{align*}
	D = \sum_{i=0}^{n}\alpha_i D_i.
\end{align*}
where $\alpha_i \in \bq$ and $\deg D_i = d_i$. Let $f_i$ be Cartier divisors such that $D_i = V(f_i)$. We shall further assume that among there exist functions $f_0,\ldots, f_{m+1}$ form a basis for degree one rational functions on $\bp^m$.

\todo{possibly change this to only have two equations}
\begin{prop}
\label{prop:pm-span-and-basis}
The functions $u^d \cdot \prod_{i=1}^n f_i^{c_i}$ so that 
\begin{align}
\label{align:pm-span}
\sum_{i=0}^{n} c_i \cdot d_i \cdot f_i = 0 & \text{ and } &c_i \geq \lfloor \alpha_i d\rfloor	
\end{align}
form a spanning set for $H^0(\bp^m, dD)$ over $k$. Furthermore, functions 
of the form $u^d \cdot \prod_{i=1}^n f_i^{c_i} $ such that
\begin{align}
\label{align:pm-basis}
c_i = -\lfloor d\alpha_i \rfloor & \text{ for } & i > m
\end{align}
form a basis for $H^0(\bp^m, dD)$ over $k$.
\end{prop}
\begin{proof}
By definition of $H^0(\bp^m,dD)$, functions satisfying conditions 
~\eqref{align:pm-span} lie in $H^0(\bp^m,dD)$. To complete the proof, it suffices to check functions satisfying conditions ~\eqref{align:pm-basis} form a basis of $H^0(\bp^m,dD)$. Note that there $\binom{m+ \lfloor dD \rfloor }{m}$ functions satisfying condition ~\eqref{align:pm-basis}. However we know $h^0(\bp^m,dD) = \binom{m+ \lfloor dD \rfloor }{m},$ so it suffices to show that those functions satisfying condition ~\eqref{align:pm-basis} are independent. This follows from the assumption that $f_0,\ldots, f_n$ form a basis of degree 1 rational functions, and so degree $\lfloor d \deg D \rfloor $ monomials in $f_0,\ldots, f_n$ form a basis of degree $\lfloor d \deg D \rfloor $ rational functions.
\end{proof}
\todo{the following bound is the generation bound beginning from evan's paper}

@@

\noindent
has a minimal presentation consisting of the $\sum_{i = 0}^{r}
{{m + i} \choose {i}}$ many generators $f_{d_i}^{\vec{v}} :=
\frac{u^{d_i} x^{\vec{v}}} {x_k^{c_i}}$ where $\vec{v} \in
\bz_{\geq 0}^{m + 1}$ such that $\sum_{j = 0}^{m} v_j = c_i$ and $z$ \todo{how many?}
many relations of the form
\[
	f_{d_i}^{\vec{v}} f_{d_j}^{\vec{w}} = f_{d_k}^{\vec{y}} f_{d_\ell}^{\vec{z}}
\]

@@

\section{Canonical Rings on Hirzebruch Surfaces}
\begin{defn}
Suppose $D = \sum_{i=1}^m \alpha_i C_i \in \mathbb{Q} \otimes \di(F_k)$.  Then define 
\begin{equation}\label{eqn:define-T=(D)}
	T_=(D) = \left\{i \in \{1, \ldots, m\}: d_i \sum_{k=1}^m d_k' \alpha_k = d_i' \sum_{k=1}^m d_k \alpha_k \right\},
\end{equation}
\begin{equation}\label{eqn:define-T+(D)}
	T_+(D) = \left\{ i \in \{1, \ldots, m\}:  \vphantom{\sum_{k=1}^m} d_i \sum_{k=1}^m \alpha_k d_k' > d_i' \sum_{k=1}^m \alpha_k d_k \right\},
\end{equation}
and
\begin{equation}\label{eqn:define-T-(D)}
	T_-(D) = \left\{ j \in \{1, \ldots, m\}: d_j \sum_{k=1}^m \alpha_k d_k' < d_j' \sum_{k=1}^m \alpha_k d_k s \right\}.
\end{equation}
\end{defn}

\begin{lem}
If $D=\sum_{i=1}^m \alpha_i C_i$ for some curve $C_i\in F_k$, such that $C_1, \ldots, C_4$ are independent curves with bi-degrees (1,0), (1,0), (0,1), and (0,1) respectively.  Then each $C_i$ corresponds to a polynomial $f_i$ where $f_1$ and $f_2$ are independent linear polynomials in $x_0$ and $x_1$ and $f_3$ and $f_4$ are independent linear polynomials in $y_0$ and $y_1$.  Then $D$ is generated in degree at most
\[
	\sum_{i\in T_=(D)} \gcd(d_i, d_i')\ell_i + \sum_{i\in T_+(D) \atop j\in T_-(D)} (d_i d_j'- d_j d_i')\ell_{i,j}
\]
 with relations in degree at most \todo{add}.
\end{lem}
\begin{proof}
Let $(d_i, d_i')$ be the bi-degrees of $f_i$.

Notice that if $\gamma \in (R_D)_d$, then 
\[
	g = \sum_{i=1}^m {f_i}^{a_i}
\]
where each $a_i \ge - \alpha_i d_i$, $\sum_{i=1}^m a_i d = 0$, and $\sum_{i=1}^m a_i d_i' = 0$.  Then we can view $g$ inside of the lattice
$S\subseteq \mathbb{R}^{n+1}$ given by ordered elements of the form $(d, a_1, \ldots, a_n)$ satisfying $a_i \ge - d \alpha_i$ for all $i$, $\sum_{i=1}^m a_i d_i = 0$, and $\sum_{i=1}^m a_i d_i' = 0$. 
From here, determining a spanning set for $(R_D)_d$ reduces to finding the extremal rays of $S$ \todo{cite previous lemma}.

To do this, we extend the method of O'Dorney \todo{cite}.  We first view the cone $T\subseteq \mathbb{R}^{n+1}$ given by elements of the form $(d, a_1, \ldots, a_n)$ satisfying $a_i \ge -\alpha_i d$ for all $i$ and $\sum_{i=1}^n a_i (d_i + d_i') =0$.  By O'Dorney, this has extremal rays given by 
\[
	e_i = (1, -\alpha_1, \ldots, -\alpha_{i-1}, \frac{\sum_{j \ne i} \alpha_j(d_j+ d_j')}{d_i + d_i'}, -\alpha_{i+1}, \ldots, - \alpha_m).
\]
Let $\Delta^{n-1}$ be the simplex with vertices $e_1, \ldots e_m$.  We can then intersect the hyperplane $H$ given $\sum_{i=1}^m  d_i X_i = 0$.  The extremal rays of $S$ are given by the set of $e_i$ contained in $H$ together with the points of intersection $e_{i,j}$ of $H$ and the line segment connecting $e_i, e_j \not \in H$.

>>>>

Write
\begin{align*}
	D = \sum_{i=0}^{n}\alpha_i D_i.
\end{align*}
where $\alpha_i \in \bq$ and $\deg D_i = a_i$. Let $f_i$ be Cartier divisors such that $D_i = V(f_i)$. We shall further assume that among there exist functions $f_0,\ldots, f_{m+1}$ form a basis for degree one rational functions on $\bp^m$.

\todo{possibly change this to only have two equations}
\begin{prop}
\label{prop:pm-span-and-basis}
The functions $u^d \cdot \prod_{i=1}^n f_i^{c_i}$ so that 
\begin{align}
\label{align:pm-span}
\sum_{i=0}^{n} c_i \cdot a_i \cdot f_i = 0 & \text{ and } &c_i \geq \lfloor \alpha_i d\rfloor	
\end{align}
form a spanning set for $H^0(\bp^m, dD)$ over $k$. Furthermore, functions 
of the form $u^d \cdot \prod_{i=1}^n f_i^{c_i} $ such that
\begin{align}
\label{align:pm-basis}
c_i = -\lfloor d\alpha_i \rfloor & \text{ for } & i > m
\end{align}
form a basis for $H^0(\bp^m, dD)$ over $k$.
\end{prop}
\begin{proof}
By definition of $H^0(\bp^m,dD)$, functions satisfying conditions 
~\eqref{align:pm-span} lie in $H^0(\bp^m,dD)$. To complete the proof, it suffices to check functions satisfying conditions ~\eqref{align:pm-basis} form a basis of $H^0(\bp^m,dD)$. Note that there $\binom{m+ \lfloor dD \rfloor }{m}$ functions satisfying condition ~\eqref{align:pm-basis}. However we know $h^0(\bp^m,dD) = \binom{m+ \lfloor dD \rfloor }{m},$ so it suffices to show that those functions satisfying condition ~\eqref{align:pm-basis} are independent. This follows from the assumption that $f_0,\ldots, f_n$ form a basis of degree 1 rational functions, and so degree $\lfloor d \deg D \rfloor $ monomials in $f_0,\ldots, f_n$ form a basis of degree $\lfloor d \deg D \rfloor $ rational functions.
\end{proof}
\todo{the following bound is the generation bound beginning from evan's paper}

@@

\noindent
has a minimal presentation consisting of the $\sum_{i = 0}^{r}
{{m + i} \choose {i}}$ generators $f_{\vec{v}} := \frac{u^{a_i}
x^{\vec{v}}}{x_k^{c_i}}$ where $\vec{v} \in \bz_{\geq 0}^{m + 1}$
such that $\sum_{j = 0}^{m} v_j = c_i$ and $z$ \todo{how many?}
relations of the form
\[
	f_{d_i}^{\vec{v}} f_{d_j}^{\vec{w}} = f_{d_k}^{\vec{y}} f_{d_\ell}^{\vec{z}}
\]

@@

\section{Canonical Rings on \texorpdfstring{$F_k$}{}}
Let $D=\sum_{i=1}^m \alpha_i C_i$ for some curve $C_i\in F_k$, such that $C_1, \ldots, C_4$ are independent curves with bi-degrees (1,0), (1,0), (0,1), and (0,1) respectively.  Let $C_i = V(f_i)$ where $f_i \in \mathscr{O}(a_i, b_i)$.  Then each $C_i$ corresponds to a polynomial $f_i$ where $f_1$ and $f_2$ are independent linear polynomials in $x_0$ and $x_1$ and $f_3$ and $f_4$ are independent linear polynomials in $y_0$ and $y_1$.
\begin{defn}
Suppose $D = \sum_{i=1}^m \alpha_i C_i \in \mathbb{Q} \otimes \di(F_k)$.  Then define 
\begin{equation}\label{eqn:define-T=(D)}
	T_=(D) = \left\{i \in \{1, \ldots, m\}: a_i \sum_{k=1}^m b_k \alpha_k = b_i \sum_{k=1}^m a_k \alpha_k \right\},
\end{equation}
\begin{equation}\label{eqn:define-T+(D)}
	T_+(D) = \left\{ i \in \{1, \ldots, m\}:  \vphantom{\sum_{k=1}^m} a_i \sum_{k=1}^m \alpha_k b_k > b_i \sum_{k=1}^m \alpha_k a_k \right\},
\end{equation}
and
\begin{equation}\label{eqn:define-T-(D)}
	T_-(D) = \left\{ j \in \{1, \ldots, m\}: a_j \sum_{k=1}^m \alpha_k b_k < b_j \sum_{k=1}^m \alpha_k a_k s \right\}.
\end{equation}
\end{defn}

\begin{lem}
$R_D$ is generated in degree at most
\[
	\sum_{i\in T_=(D)} \gcd(a_i, b_i)\ell_i + \sum_{i\in T_+(D) \atop j\in T_-(D)} (a_i b_j- a_j b_i)\ell_{i,j}.
\]
\end{lem}
\begin{proof}
Let $(a_i, b_i)$ be the bi-degrees of $f_i$.

Notice that if $\gamma \in (R_D)_d$, then 
\[
	g = \sum_{i=1}^m {f_i}^{c_i}
\]
where each $c_i \ge - \alpha_i c_i$, $\sum_{i=1}^m c_i a_i = 0$, and $\sum_{i=1}^m c_i b_i = 0$.  Then we can view $g$ inside of the lattice
$S\subseteq \mathbb{R}^{n+1}$ given by ordered elements of the form $(d, c_1, \ldots, c_n)$ satisfying $c_i \ge - d \alpha_i$ for all $i$, $\sum_{i=1}^m c_i a_i = 0$, and $\sum_{i=1}^m c_i b_i = 0$. 
From here, determining a spanning set for $(R_D)_d$ reduces to finding the extremal rays of $S$ \todo{cite previous lemma}.

To do this, we extend the method of O'Dorney \todo{cite}.  We first view the cone $T\subseteq \mathbb{R}^{n+1}$ given by elements of the form $(d, c_1, \ldots, c_n)$ satisfying $c_i \ge -\alpha_i d$ for all $i$ and $\sum_{i=1}^n c_i (a_i + b_i) =0$.  By O'Dorney, this has extremal rays given by 
\[
	e_i = (1, -\alpha_1, \ldots, -\alpha_{i-1}, \frac{\sum_{j \ne i} \alpha_j(a_j+ b_j)}{a_i + b_i}, -\alpha_{i+1}, \ldots, - \alpha_m).
\]
Let $\Delta^{n-1}$ be the simplex with vertices $e_1, \ldots e_m$.  We can then intersect the hyperplane $H$ given $\sum_{i=1}^m  a_i X_i = 0$.  The extremal rays of $S$ are given by the set of $e_i$ contained in $H$ together with the points of intersection $e_{i,j}$ of $H$ and the line segment connecting $e_i, e_j \not \in H$.
